Joint estimation of ancestral states and phylogenetic model parameters is inconsistent

### 2017-12-07 joint_inf.tex
Going through main text again :+1:

\begin{abstract}
%EM This abstract is really nice! :+1:
Maximum likelihood is a widespread and well-understood approach to estimate the phylogenetic parameters of topology and branch length given observed character data.
Since a phylogeny has unknown ancestral states, computing a maximum likelihood solution becomes a problem involving latent variables.
Classically, these unknown intermediate states are averaged out, which leads to consistent estimation of the topology and continuous model parameters.
Recently, a computationally-efficient approach has been proposed to jointly maximize over these unknown states and phylogenetic parameters.
%EM I think that should either use "joint" or "profile" just to have a single word to describe things for most of the text, except when we connect it to the stats literature.
While this method of joint maximum likelihood estimation can obtain estimates more efficiently, it has the potential to result in undesired properties of the estimates.
We show that, like other methods of maximum likelihood where the number of unknowns grows with observations, this method of jointly estimating phylogenetic parameters along with ancestral states is not consistent in general.
There are a number of cases where generating under a four-taxon tree results in a joint estimation procedure converging on the wrong topology.
If the topology is known a priori, there are similar cases where the branch lengths are consistently overestimated in the limit of infinite data.
Although joint maximum likelihood may turn out to be a useful tool in practice, care should be taken when using this approach.
\end{abstract}

@@

Existing consistency proofs for maximum likelihood phylogenetics \cite{RoyChoudhury2015-ta} apply only to estimating model parameters when the ancestral sequences have been integrated out of the likelihood.
These proofs do not readily extend to include estimating ancestral states.
Moreover, examples of inconsistency arising from including a large number of additional parameters \cite{Neyman1948-tt} indicate that joint inference of trees and ancestral states may not enjoy good statistical properties.
Although \cite{Sagulenko2017-jo} explicitly warn that their approximation is for the case of branch lengths $\ll 1$, their work motivates understanding when exactly this approximate technique breaks down, even in the limit of infinite sequence data.

In this paper, we show that the joint inference of trees and ancestral sequences is not consistent in general.
To do so, we use a binary symmetric model and imagine data being generated on the four-taxon ``Farris zone'' \cite{Siddall1998-hq} tree, and we construct bounds on the joint likelihood to demarcate a sizeable area of long branch lengths in which joint inference is guaranteed to give the wrong tree in the limit of infinite data.
We find similar areas where joint inference consistently overestimates interior branch lengths when the topology is known.

@@

We parameterize branches on the unique unrooted four-tip phylogenetic tree in ways known as the ``Farris'' and ``Felsenstein'' trees (Fig.~\ref{fig:farris-fels-top}).
In the classical configuration of each of these trees, the interior branch length parameters are equal to the bottom two parameters.
We show in the Appendix that, in the case of the Farris tree, fixing both the top two branch parameters to be equal and the bottom two branch parameters to be equal will obtain the same maximum likelihood estimate as in the case of arbitrary branch parameters.

\begin{figure}
\centering
\begin{subfigure}{.45\linewidth}
\centering
\includegraphics[width=\textwidth]{farris_blank}
\caption[short]{Farris tree $\tau_1$}
\end{subfigure}
\begin{subfigure}{.45\linewidth}
\centering
\includegraphics[width=\textwidth]{felsenstein_blank}
\caption[short]{Felsenstein tree $\tau_2$}
\end{subfigure}
\caption{Two four-taxon trees}
\label{fig:farris-fels-top}
\end{figure}

@@

Fidelities have useful algebraic properties, and generating probabilities using the Hadamard transform have an especially simple form (see \eqref{eq:hadamard_probability} in the Appendix).
For a four-taxon tree, define the general branch fidelity parameter $t=\{\theta_1,\theta_2,\theta_3,\theta_4,\theta_5\}$ where fidelities are ordered in the order of the taxa with the internal branch last.
The trees in Fig.~\ref{fig:farris-fels-top} have $\theta_1=\theta_3=x$, $\theta_2=\theta_4=y$, and $\theta_5=w$.

@@

The typical approach to is to maximize the likelihood marginalized across ancestral states
\begin{equation}
\label{eq:marginal_likelihood}
\tilde{L}_\nCols(\tau, t; \fullAlignment) = \sum_{\fullAncestralStates\in\alphabet^{\nAncestralStateRows\times\nCols}} \ L_\nCols(\tau, t; \fullAlignment, \fullAncestralStates)
\end{equation}
to estimate the tree $\tau$ and branch fidelities $t$.

@@

Assume $\fullAlignment$ is generated from topology $\tau^*$ and branch fidelities $t^*$.
Use $\ell_{\tau^*,t^*}(\tau, t)$ to denote the expected per-site log likelihood, which can be thought of as the infinite-length sequence limit
$$
\frac{1}{n}\log L_\nCols'(\tau, t; \fullAlignment) \rightarrow \ell_{\tau^*,t^*}(\tau, t).
$$
We give $\ell$ explicitly in \eqref{eq:site_pattern_profile_likelihood_mean} in the Appendix.

@@

To show an inconsistency in topology estimation, we start with true generating parameters $t^*=\{x^*, y^*, x^*, y^*, y^*\}$ on the Farris topology.
We show that, as $\nCols\rightarrow\infty$, there exist values for $x^*$ and $y^*$ such that the value of the likelihood after maximizing using joint inference is greater for the Felsenstein topology than for the true, generating Farris topology.
To do so, we construct an upper bound for the likelihood given the Farris topology as a function of $x^*$ and $y^*$, and, similarly, we construct a lower bound for the likelihood given the Felsenstein topology.
The values of $x^*$ and $y^*$ where the likelihood in the Felsenstein case is larger than the likelihood in the Farris case indicate an inconsistency (Fig.~\ref{fig:inconsistency-farris}).
\begin{restatable}{theorem}{topoInconsist}
%EM Wow this is hard to parse!
%EM 1. The main message is that there is a region on which the wrong topology is preferred. This should be stated in words before the theorem, with the link to the figure.
%EM 2. To see that in the theorem, we have to wind our way through inequalities facing different ways. Can we just say that there is a region? What do these L's tell us?
%EM 3. L_0 and L_1 are totally different that L_n, but with very similar notation.
%das: is this better, or did you want to get rid of the upper/lower bound constants entirely? I have them since they are a closed form function of x* and y* so it's clear that all we need is a region where some f(x*, y*) > 0. Are you saying the paragraph I just put before the theorem is a better statement of the theorem?
Let $t^*=\{x^*, y^*, x^*, y^*, y^*\}$ and $t=\{x, y, x, y, w\}$.
There exist $C_0(x^*, y^*), C_1(x^*, y^*)$ and $0 < x^*, y^* < 1$ such that
$$
\max_{t} \ \ell_{\tau_1,t^*}(\tau_1, t) \le C_0(x^*, y^*),
$$
$$
\max_{t} \ \ell_{\tau_1,t^*}(\tau_2, t) \ge C_1(x^*, y^*)
$$
with $C_1(x^*, y^*) > C_0(x^*, y^*)$.
\end{restatable}
Fig.~\ref{fig:inconsistency-farris} shows the values of $x^*$ and $y^*$ where $C_1(x^*, y^*) > C_0(x^*, y^*)$.
Intuitively, $\tau_2$ is favored in performing joint maximum likelihood since the likelihood for $\tau_2$ has more ``degrees of freedom''---Table~\ref{tab:likelihoods} shows that $\tau_1$ has only three possible forms for its likelihood while $\tau_2$ has many more.
This results in more possible maxima for $\tau_2$ even if data are generated from $\tau_1$.

\begin{figure}
\centering
% obtained by running
% python joint_inf_plot.py --analytic --topology --delta .001 --plot-name figures/topology-inconsistency.svg
\includegraphics[width=\textwidth]{topology-inconsistency-inkscape}
\caption{
    Regions of inconsistency.
    Due to the looseness of the upper and lower bounds, the parameters in the upper right do not necessarily indicate consistency, though all parameters in the labeled region result in an inconsistency.
}
\label{fig:inconsistency-farris}
\end{figure}

@@

We now consider the problem of branch length estimation on the correct tree using joint ancestral state and branch length estimation.
As described above, we will use the equivalent but different notion of branch fidelities.
We analyze two settings on the Farris tree, corresponding to whether some branch fidelities are fixed at their true values or not.
As above, assume that data is generated from the ``Farris zone'' with two top branches of fidelity $x^*$ and all other branches of fidelity $y^*$.
In the first ``restricted'' case, we show that for a nontrivial subset of possible values for $x^*$ and $y^*$, the interior branch fidelity parameter $w$ will be consistently overestimated as exactly equal to one instead of its true value of $y^*$.
That is, if we know $x^*$ and $y^*$, then for
$$
\hat{w} = \arg\max_{w} \ \ell_{\tau_1,t^*}(\tau_1, \{x^*,y^*,x^*,y^*,w\}),
$$
we show there is a set of values for $x^*$ and $y^*$ where $\hat{w}\equiv 1$.
In the general case, we do the same but with
$$
(\hat{x}, \hat{y}, \hat{w}) = \arg\max_{x,y,w} \ \ell_{\tau_1,t^*}(\tau_1, \{x,y,x,y,w\}).
$$
Again, we find a region for which the inferred values do not converge to the generating values.
These situations are in contrast to the approach using marginal likelihood where necessarily $\hat{w}$ converges to $y^*$ as the number of observations grows.

@@

Fix $x=x^*$ and $y=y^*$ to their true, generating values and proceed to estimate the internal branch parameter $w$.
\begin{restatable}{theorem}{restrictedBranchInconsist}
\label{thm:restricted-bl}
Let
$$
\alpha_1 := \alpha_1(x^*, y^*) = \frac{1}{8} \left(1+(x^*)^2+(y^*)^2+4x^*(y^*)^2+(x^*)^2(y^*)^2\right),
$$
$$
\alpha_2 := \alpha_2(x^*, y^*) = \frac{1}{8}\left(1+(x^*)^2+(y^*)^2-4x^*(y^*)^2+(x^*)^2(y^*)^2\right),
$$
$$
\beta := \beta(x^*, y^*) = 1+(x^*)^2+(y^*)^2+(x^*)^2(y^*)^2,
$$
$$
\gamma := \gamma(x^*, y^*) = 4x^*y^*.
$$
The maximum likelihood value $\hat{w} = 1$ if
$$
\beta^2-\gamma^2-2\gamma^2\alpha_1-2\gamma^2\alpha_2+2\gamma\alpha_1\beta-2\gamma\alpha_2\beta \ge 0.
$$
\end{restatable}
The inequality in Theorem~\ref{thm:restricted-bl} is a function of the generating parameters $x^*$ and $y^*$.
This region is shown in Fig.~\ref{fig:bl-inconsistency}.
Intuitively, since we perform maximization over internal states, branch fidelities will have a tendency to be overestimated.
Seeing no change along a branch more likely increases the likelihood by introducing a term of $(1+\theta)$ instead of $(1-\theta)$, and branch fidelities will be positively biased due to this.

\begin{figure}
\centering
% obtained by running
% python joint_inf_plot.py --analytic --restricted-branch-lengths --delta .001 --plot-name figures/branch-length-inconsistency.svg
\includegraphics[width=\textwidth]{branch-length-inconsistency-inkscape}
\caption{Region of inconsistency with $x=x^*$ and $y=y^*$ fixed.}
\label{fig:bl-inconsistency}
\end{figure}

@@

The general case is more challenging to analyze and so we will get somewhat weaker bounds.
In the general case, $\hat{w}$ is a function of $x^*$, $y^*$, $\hat{x}$, and $\hat{y}$.
%EM This is the first we're hearing of the quadratic formula in this version.
Looking to the previous section, the region where $\hat{w}=1$ will still be given by the inequality in Theorem~\ref{thm:restricted-bl}, only with $\gamma$ and $\beta$ now being functions of $\hat{x}$ and $\hat{y}$ instead of $x^*$ and $y^*$.
Assume we know $\hat{x}$ and $\hat{y}$ as functions of $x^*$ and $y^*$ only.
%EM Here would be a good place to introduce the concept that youre going to be saying that either the pendant branch lengths are bad, or what is going to be 1.
%EM differing notions of "bad" will lead to different theorems.
We show that there are similar bounds as in the restricted case, though we need to take into account the unknown values of $\hat{x}$ and $\hat{y}$.
We fix bounds on these estimates and show that, in the general case, we either estimate $\hat{w}$ to be one or fall outside of the specified bounds for $\hat{x}$ or $\hat{y}$, indicating a poor estimate in at least one of the three unknown branch parameters.
\begin{restatable}{theorem}{generalBranchInconsist}
\label{thm:general-bl}
For
$$
\beta := \beta(x^*, y^*) = 1+(x^*)^2+(y^*)^2+(x^*)^2(y^*)^2,
$$
$$
\gamma := \gamma(x^*, y^*) = 4x^*y^*,
$$
and bounds
$$
\gamma_L := \gamma_{L}(x^*, y^*) \le \gamma(\hat{x}, \hat{y}),
$$
$$
\gamma_U := \gamma_{U}(x^*, y^*) \ge \gamma(\hat{x}, \hat{y}),
$$
and
$$
\beta_L := \beta_{L}(x^*, y^*) \le \beta(\hat{x}, \hat{y}),
$$
the maximum likelihood value $\hat{w}=1$ when
$$
-\gamma_{U}^2\left(1 + \frac{1}{2}\beta\right) + 2\gamma_{L}\beta_{L}x^*(y^*)^2 + \beta_{L}^2 \ge 0.
$$
\end{restatable}
Similar to Theorem~\ref{thm:restricted-bl}, the inequality in Theorem~\ref{thm:general-bl} is also a function of only $x^*$ and $y^*$, though general forms for $\gamma_L$, $\gamma_U$, and $\beta_L$ are difficult to obtain.
%EM I think you put this loosest bound in for exposition's sake, but I don't think it helps clarify.
%EM On the other hand, I think that explaining that the restricted case fits in this framework is great.
The tightest possible bounds are $\gamma_{L} = \gamma_{U} = \gamma$ and $\beta_{L} = \beta$, which is the restricted case of the previous section and given in Fig.~\ref{fig:bl-inconsistency}.
For an intermediate bound, define a specified level of error in estimates for $\hat{x}$ and $\hat{y}$ so that
$$
x^*-x_{L} \le \hat{x} \le x^*+x_{U}
$$
and
$$
y^*-y_{L} \le \hat{y} \le y^*+y_{U}.
$$
Then $\gamma_L, \gamma_U$ and $\beta_L$ are defined directly from the bounds on $\hat{x}$ and $\hat{y}$.
As an example, the region to the left side of the curve in Fig.~\ref{fig:bl-loose-inconsistency} shows the case where $x_L=x_U=y_L=y_U=0.1$.
In this case, we will either estimate $\hat{w}$ to be one or estimate either $\hat{x}$ or $\hat{y}$ to be more than $0.1$ away from its true value.

\begin{figure}
\centering
% obtained by running
% python joint_inf_plot.py --analytic --general-branch-lengths --delta .001 --plot-name figures/bl-loose-inconsistency.svg
\includegraphics[width=\textwidth]{bl-loose-inconsistency-inkscape}
%EM If you read this caption without reading the text, would you be able to understand what's going on? I think I'd be confused because in the caption you are stating something that looks good, and then the figure describes xhat and yhat as being poorly estimated.
\caption{
    Region of inconsistency.
    In the marked region, either $\hat{w}=1$ or one of $x^*-0.1 \le \hat{x} \le x^*+0.1$ or $y^*-0.1 \le \hat{y} \le y^*+0.1$ will not be true, resulting in poor estimation of the noninterior branch parameters.
}
\label{fig:bl-loose-inconsistency}
\end{figure}

@@

The region of inconsistency in Fig.~\ref{fig:bl-loose-inconsistency} is a conservative estimate of where $\hat{w}=1$.
To determine how conservative, we use the method of basin-hopping \cite{Wales1997} to perform joint maximum likelihood estimation using numerical optimization, described in the Appendix.
Since empirical estimates can be unstable, we perform two maximizations---one over $0 \le x,y,w \le 1$ and one over $0 \le x,y \le 1$ with $w=1$---and take the value of $\hat{w}$ with the higher likelihood.
We compute these maxima over a lattice in steps of $10^{-2}$ for $x^*$ and $y^*$ from $10^{-2}$ to $1-10^{-2}$ (Fig.~\ref{fig:bl-general-inconsistency}): we do not include zero or one in our lattice to further stabilize the fits, as these cases can result in pathologies.
%EM Noo, I need to make a name for this sort of sentence-- the figure announcement sentence? Sentences should read "Here is a statement (Fig. X)."
%EM There is a big bomb to drop here, which is that what is 1 all over the place!
%EM Question: it seems to me that even in the non-white region, the BL inferences aren't correct. Is that true? We should say so if. I don't want to make extra work for you, but a figure in which you take the difference between the correct value and the inferred value might be nice.
%das: it seemed like the estimates were decent to me, but I can do this super easily (output is pickled). Though are you talking about another plot or just a check that we can add a sentence like "even in cases where \hat{w} is not one w is overestimated/poorly estimated."

@@

We have shown that the method of joint inference when estimating phylogenetic parameters is not consistent in general.
Specifically, in the case of four-taxon trees with infinite data, we have obtained nontrivial regions of generating parameters that result in two types of inconsistency: first, inferring the incorrect topology and, second, estimating severely biased branch lengths even given the correct topology.
In all cases, these regions of inconsistency arise when the branches of the generating trees are ``long,'' that is, when branch fidelities are closer to zero than to one.

The method of profile likelihood is well-studied \cite{Murphy2000-ry}.
Many properties regarding the performance of maximum likelihood estimates obtained using this approach are known, and many methods exist to overcome their undesirable properties, e.g., \cite{Geman1982}.
Joint inference of phylogenetic parameters and ancestral states \cite{Sagulenko2017-jo} is a similar approach to profile likelihood.
Elsewhere in statistics literature, the failure of maximum likelihood estimates to obtain consistent estimates in the case of infinite parameters have been shown by the Neyman-Scott paradox \cite{Neyman1948-tt}, though parameters tending to infinity is not a necessary condition for inconsistency \cite{LeCam1990}.
Until recently \cite{RoyChoudhury2015-ta}, results on the properties of standard maximum likelihood estimates of phylogeny as in \eqref{eq:marginal_likelihood} have been difficult to prove, and no results have been obtained for profile likelihood.
We have furthered progress in understanding the limitations of this joint maximum likelihood approach.

>>>>

\begin{abstract}
Maximum likelihood is a widespread and well-understood approach to estimate the phylogenetic parameters of topology and branch length given observed character data.
Since a phylogeny has unknown ancestral states, computing a maximum likelihood solution becomes a problem involving latent variables.
Classically, these unknown intermediate states are averaged out, which leads to consistent estimation of the topology and continuous model parameters.
Recently, a computationally-efficient approach has been proposed to jointly maximize over these unknown states and phylogenetic parameters.
While this method of joint maximum likelihood estimation can obtain estimates more efficiently, it has the potential to result in undesired properties of the estimates.
We show that, like other methods of maximum likelihood where the number of unknowns grows with observations, this method of jointly estimating phylogenetic parameters along with ancestral states is not consistent in general.
There are a number of cases where generating under a four-taxon tree results in a joint estimation procedure converging on the wrong topology.
If the topology is known a priori, there are similar cases where the branch lengths are consistently overestimated in the limit of infinite data.
Although joint maximum likelihood may turn out to be a useful tool in practice, care should be taken when using this approach.
\end{abstract}

@@

Existing consistency proofs for maximum likelihood phylogenetics \cite{RoyChoudhury2015-ta} apply only to estimating model parameters when the ancestral sequences have been integrated out of the likelihood.
These proofs do not readily extend to include estimating ancestral states.
Moreover, examples of inconsistency arising from including a large number of additional parameters \cite{Neyman1948-tt} indicate that joint inference of trees and ancestral states may not enjoy good statistical properties.
In this case those additional parameters come in the form of the bases at ancestral sequences.
Although \cite{Sagulenko2017-jo} explicitly warn that their approximation is for the case of branch lengths $\ll 1$, their work motivates understanding when exactly this approximate technique breaks down, even in the limit of infinite sequence data.

In this paper, we show that the joint inference of trees and ancestral sequences is not consistent in general.
To do so, we use a binary symmetric model and imagine data being generated on the four-taxon ``Farris zone'' \cite{Siddall1998-hq} tree, and we construct bounds on the joint likelihood to demarcate a sizeable area of long branch lengths in which joint inference is guaranteed to give the wrong tree in the limit of infinite data.
We find similar areas where joint inference consistently overestimates interior branch lengths when the topology is known and fixed.

@@

We parameterize branches on the unique unrooted four-tip phylogenetic tree in ways known as the ``Farris'' and ``Felsenstein'' trees (Fig.~\ref{fig:farris-fels-top}).
In the classical configuration of each of these trees, the interior branch length parameters are equal to the bottom two parameters.
We show in the Appendix that, in the case of the Farris tree, doing inference fixing both the top two branch parameters to be equal and the bottom two branch parameters to be equal will obtain the same maximum likelihood estimate as in the case of arbitrary branch parameters.

\begin{figure}
\centering
\begin{subfigure}{.45\linewidth}
\centering
\includegraphics[width=\textwidth]{farris_blank}
\caption[short]{Farris tree $\tau_1$}
\end{subfigure}
\begin{subfigure}{.45\linewidth}
\centering
\includegraphics[width=\textwidth]{felsenstein_blank}
\caption[short]{Felsenstein tree $\tau_2$}
\end{subfigure}
\caption{Two four-taxon trees with fidelities as labeled: $\theta_1=\theta_3=x$, $\theta_2=\theta_4=y$, and $\theta_5=w$.
}
\label{fig:farris-fels-top}
\end{figure}

@@

Fidelities have useful algebraic properties, and generating probabilities using the Hadamard transform have an especially simple form (see \eqref{eq:hadamard_probability} in the Appendix).
For a four-taxon tree, define the general branch fidelity parameter $t=\{\theta_1,\theta_2,\theta_3,\theta_4,\theta_5\}$ where fidelities are ordered in the order of the taxa with the internal branch last (Fig.~\ref{fig:farris-fels-top}).

@@

The classical approach to is to maximize the likelihood marginalized across ancestral states
\begin{equation}
\label{eq:marginal_likelihood}
\tilde{L}_\nCols(\tau, t; \fullAlignment) = \sum_{\fullAncestralStates\in\alphabet^{\nAncestralStateRows\times\nCols}} \ L_\nCols(\tau, t; \fullAlignment, \fullAncestralStates)
\end{equation}
to estimate the tree $\tau$ and branch fidelities $t$.

@@

Assume $\fullAlignment$ is generated from topology $\tau^*$ and branch fidelities $t^*$.
Use $\ell_{\tau^*,t^*}(\tau, t)$ to denote the expected per-site log likelihood, which can be thought of as the infinite-length sequence limit
$$
\frac{1}{n}\log L_\nCols'(\tau, t; \fullAlignment) \rightarrow \ell_{\tau^*,t^*}(\tau, t).
$$
We give $\ell$ explicitly as \eqref{eq:site_pattern_profile_likelihood_mean} in the Appendix.

@@

To show an inconsistency in topology estimation, we start with true generating parameters $t^*=\{x^*, y^*, x^*, y^*, y^*\}$ on the Farris topology.
We show that, as $\nCols\rightarrow\infty$, there exist values for $x^*$ and $y^*$ such that the value of the likelihood after maximizing using joint inference is greater for the Felsenstein topology than for the true, generating Farris topology.
To do so, we construct an upper bound $C_0(x^*, y^*)$ for the likelihood given the Farris topology as a function of $x^*$ and $y^*$, and, similarly, we construct a lower bound $C_1(x^*, y^*)$ for the likelihood given the Felsenstein topology.
When $C_0(x^*, y^*) < C_1(x^*, y^*)$, the likelihood in the Felsenstein case is larger than the likelihood in the Farris case, demonstrating inconsistency (Fig.~\ref{fig:inconsistency-farris}).
\begin{restatable}{theorem}{topoInconsist}
Let $t^*=\{x^*, y^*, x^*, y^*, y^*\}$ and $t=\{x, y, x, y, w\}$.
There exist $C_0(x^*, y^*), C_1(x^*, y^*)$ and a set of $0 < x^*, y^* < 1$ such that
$$
\max_{t} \ \ell_{\tau_1,t^*}(\tau_1, t) \le C_0(x^*, y^*),
$$
$$
C_1(x^*, y^*) \le \max_{t} \ \ell_{\tau_1,t^*}(\tau_2, t)
$$
with $C_0(x^*, y^*) < C_1(x^*, y^*)$.
\end{restatable}
The proof of this theorem is by a detailed examination of inequalities.
The only intuition we can offer is that $\tau_2$ is favored in performing joint maximum likelihood since the likelihood for $\tau_2$ has more ``degrees of freedom''---Table~\ref{tab:likelihoods} shows that $\tau_1$ has only three possible forms for its likelihood while $\tau_2$ has many more.
This enables more possible maxima for $\tau_2$ even if data are generated from $\tau_1$.

\begin{figure}
\centering
% obtained by running
% python joint_inf_plot.py --analytic --topology --delta .001 --plot-name figures/topology-inconsistency.svg
\includegraphics[width=\textwidth]{topology-inconsistency-inkscape}
\caption{
    Regions of topological inconsistency in terms of fidelities.
    Due to the looseness of the upper and lower bounds, the parameters in the upper right do not necessarily indicate consistency, though all parameters in the labeled region result in an inconsistency.
}
\label{fig:inconsistency-farris}
\end{figure}

@@

We now consider the problem of branch length estimation on the correct tree using joint ancestral state and branch length estimation.
As described above, we will use the equivalent but different notion of branch fidelities.
We analyze two settings on the Farris tree, corresponding to whether some branch fidelities are fixed at their true values or not.
As above, assume that data is generated from the ``Farris zone'' with two top branches of fidelity $x^*$ and all other branches of fidelity $y^*$.
In the first ``restricted'' case, we show that for a nontrivial subset of possible values for $x^*$ and $y^*$, the interior branch fidelity parameter $w$ will be consistently overestimated as exactly equal to one (zero branch length) instead of its true value of $y^*$.
That is, if we know $x^*$ and $y^*$, then for
$$
\hat{w} = \arg\max_{w} \ \ell_{\tau_1,t^*}(\tau_1, \{x^*,y^*,x^*,y^*,w\}),
$$
we show there is a set of values for $x^*$ and $y^*$ where $\hat{w}\equiv 1$.
In the general case, we do the same but with
$$
(\hat{x}, \hat{y}, \hat{w}) = \arg\max_{x,y,w} \ \ell_{\tau_1,t^*}(\tau_1, \{x,y,x,y,w\}).
$$
Again, we find a region for which the inferred values do not converge to the generating values.
These situations are in contrast to the approach using marginal likelihood where $\hat{w}$ necessarily converges to $y^*$ as the number of observations grows.

@@

Fix estimated fidelities $x=x^*$ and $y=y^*$ to their true, generating values and proceed to estimate the internal branch parameter $w$.
\begin{restatable}{theorem}{restrictedBranchInconsist}
\label{thm:restricted-bl}
Let
$$
\alpha_1 := \alpha_1(x^*, y^*) = \frac{1}{8} \left(1+(x^*)^2+(y^*)^2+4x^*(y^*)^2+(x^*)^2(y^*)^2\right),
$$
$$
\alpha_2 := \alpha_2(x^*, y^*) = \frac{1}{8}\left(1+(x^*)^2+(y^*)^2-4x^*(y^*)^2+(x^*)^2(y^*)^2\right),
$$
$$
\beta := \beta(x^*, y^*) = 1+(x^*)^2+(y^*)^2+(x^*)^2(y^*)^2,
$$
$$
\gamma := \gamma(x^*, y^*) = 4x^*y^*.
$$
The maximum likelihood value $\hat{w}$ is equal to 1 if
$$
\beta^2-\gamma^2-2\gamma^2\alpha_1-2\gamma^2\alpha_2+2\gamma\alpha_1\beta-2\gamma\alpha_2\beta \ge 0.
$$
\end{restatable}
The inequality in Theorem~\ref{thm:restricted-bl} is a function of the generating parameters $x^*$ and $y^*$.
This region is shown in Fig.~\ref{fig:bl-inconsistency}.
Intuitively, since we perform maximization over internal states, branch fidelities will have a tendency to be overestimated.
Seeing no change along a branch more likely increases the likelihood by introducing a term of $(1+\theta)$ instead of $(1-\theta)$, and branch fidelities will be positively biased due to this.

\begin{figure}
\centering
% obtained by running
% python joint_inf_plot.py --analytic --restricted-branch-lengths --delta .001 --plot-name figures/branch-length-inconsistency.svg
\includegraphics[width=\textwidth]{branch-length-inconsistency-inkscape}
\caption{Region of branch parameter inconsistency in terms of fidelities $x=x^*$ and $y=y^*$ that are fixed to their correct values.}
\label{fig:bl-inconsistency}
\end{figure}

@@

The general case is more challenging to analyze and so we will get somewhat weaker bounds.
In the general case, $\hat{w}$ is a function of $x^*$, $y^*$, $\hat{x}$, and $\hat{y}$.
Looking to the previous section, the region where $\hat{w}=1$ will still be given by the inequality in Theorem~\ref{thm:restricted-bl}, only with $\gamma$ and $\beta$ now being functions of $\hat{x}$ and $\hat{y}$ instead of $x^*$ and $y^*$.
Assume we know $\hat{x}$ and $\hat{y}$ as functions of $x^*$ and $y^*$ only.
We show that there are similar bounds as in the restricted case, though we need to take into account the unknown values of $\hat{x}$ and $\hat{y}$.
We fix bounds on these estimates and show that, in the general case, we either estimate $\hat{w}$ to be one or fall outside of specified bounds for $\hat{x}$ or $\hat{y}$, indicating a poor estimate in at least one of the three unknown branch parameters.
\begin{restatable}{theorem}{generalBranchInconsist}
\label{thm:general-bl}
Define $\gamma(x, y) = 4xy$.
For
$$
\beta := \beta(x^*, y^*) = 1+(x^*)^2+(y^*)^2+(x^*)^2(y^*)^2,
$$
$$
\gamma := \gamma(x^*, y^*),
$$
bounds
$$
\gamma_L := \gamma_{L}(x^*, y^*) \le \gamma(\hat{x}, \hat{y}),
$$
$$
\gamma_U := \gamma_{U}(x^*, y^*) \ge \gamma(\hat{x}, \hat{y}),
$$
and
$$
\beta_L := \beta_{L}(x^*, y^*) \le \beta(\hat{x}, \hat{y}),
$$
the maximum likelihood value $\hat{w}=1$ when
$$
-\gamma_{U}^2\left(1 + \frac{1}{2}\beta\right) + 2\gamma_{L}\beta_{L}x^*(y^*)^2 + \beta_{L}^2 \ge 0.
$$
\end{restatable}
Similar to Theorem~\ref{thm:restricted-bl}, the inequality in Theorem~\ref{thm:general-bl} is also a function of only $x^*$ and $y^*$.
Although general forms for $\gamma_L$, $\gamma_U$, and $\beta_L$ are difficult to obtain, we can make progress by bounding the error in branch length estimation for the pendant or the central branch.
If we don't want to allow any error in pendant branches, we can use the tightest possible bounds $\gamma_{L} = \gamma_{U} = \gamma$ and $\beta_{L} = \beta$, which is the restricted case of the previous section and given in Fig.~\ref{fig:bl-inconsistency}.
For an intermediate bound, define a specified allowable level of error in estimates for $\hat{x}$ and $\hat{y}$ so that
$$
x^*-x_{L} \le \hat{x} \le x^*+x_{U}
$$
and
$$
y^*-y_{L} \le \hat{y} \le y^*+y_{U}.
$$
Then $\gamma_L, \gamma_U$ and $\beta_L$ are defined directly from the bounds on $\hat{x}$ and $\hat{y}$.
As an example, the region to the left side of the curve in Fig.~\ref{fig:bl-loose-inconsistency} shows the case where $x_L=x_U=y_L=y_U=0.1$.
In this case, joint inference will either estimate $\hat{w}$ to be one or estimate either $\hat{x}$ or $\hat{y}$ to be more than $0.1$ away from its true value.

\begin{figure}
\centering
% obtained by running
% python joint_inf_plot.py --analytic --general-branch-lengths --delta .001 --plot-name figures/bl-loose-inconsistency.svg
\includegraphics[width=\textwidth]{bl-loose-inconsistency-inkscape}
\caption{
    Region of branch parameter inconsistency in terms of fidelities.
    In the marked region, either $\hat{w}=1$ or one of $x^*-0.1 \le \hat{x} \le x^*+0.1$ or $y^*-0.1 \le \hat{y} \le y^*+0.1$ will not be true, resulting in poor estimation of the noninterior branch parameters.
}
\label{fig:bl-loose-inconsistency}
\end{figure}

@@

Direct numerical optimization confirms our theoretically-derived bounds and provides more detailed picture.
The region of inconsistency in Fig.~\ref{fig:bl-loose-inconsistency} is a conservative estimate of where $\hat{w}=1$.
To determine how conservative, we use the method of basin-hopping \cite{Wales1997} to perform joint maximum likelihood estimation using numerical optimization, as described in the Appendix.
%EM This could be considered a little unfair. Optimization of 3 variables is harder than optimizing two, so when there is an optimization problem with the 3-variable case then w=1 has an advantage. I don't have a suggestion to change things.
Since empirical estimates can be unstable, we perform two maximizations---one over $0 \le x,y,w \le 1$ and one over $0 \le x,y \le 1$ with $w=1$---and take the value of $\hat{w}$ with the higher likelihood.
We compute these maxima over a lattice in steps of $10^{-2}$ for $x^*$ and $y^*$ from $10^{-2}$ to $1-10^{-2}$ (Fig.~\ref{fig:bl-general-inconsistency}): we do not include zero or one in our lattice to further stabilize the fits, as these cases can result in pathologies.
%EM Question: it seems to me that even in the non-white region, the BL inferences aren't correct. Is that true? We should say so if. I don't want to make extra work for you, but a figure in which you take the difference between the correct value and the inferred value might be nice.
%das: it seemed like the estimates were decent to me, but I can do this super easily (output is pickled). Though are you talking about another plot or just a check that we can add a sentence like "even in cases where \hat{w} is not one w is overestimated/poorly estimated."
%EM I'm pretty sure I can see that there's bias in the estimates even in the non-white area. A plot of the difference between the two would be cool. Can we zoom into the region with x* and y* greater than 0.6 so that the scale doesn't get thrown off by big differences where x* and y* are zero?

@@

We have shown that the method of joint inference when estimating phylogenetic parameters \cite{Sagulenko2017-jo} is not consistent in general.
Specifically, in the case of four-taxon trees with infinite data, we have obtained nontrivial regions of generating parameters that result in two types of inconsistency: first, where joint inference infers the incorrect topology and, second, where it estimates severely biased branch lengths even given the correct topology.
In all cases, these regions of inconsistency arise when the branches of the generating trees are ``long,'' that is, when branch fidelities are closer to zero than to one.

Joint inference of tree parameters and ancestral sequences is a type of profile likelihood, a well-studied subject in statistics \cite{Murphy2000-ry}.
Many properties regarding the performance of maximum likelihood estimates obtained using this approach are known, and many methods exist to overcome their undesirable properties, e.g., \cite{Geman1982}.
Elsewhere in statistics literature, the failure of maximum likelihood estimates to obtain consistent estimates as the number of parameters goes to infinity have been shown by the Neyman-Scott paradox \cite{Neyman1948-tt}, though parameters tending to infinity is not a necessary condition for inconsistency \cite{LeCam1990}.
Until recently \cite{RoyChoudhury2015-ta}, results on the properties of standard maximum likelihood estimates of phylogeny as in \eqref{eq:marginal_likelihood} have been difficult to prove, and no results have been obtained for profile likelihood.
We have furthered progress in understanding the limitations of this joint maximum likelihood approach.
